Make a README

### 2025-03-05 code.qmd
Rework README
Rework the instructions for creating
a README to better fit to papers/
manuscripts, rather than to software
projects.

>>>>

::: {#wrn-dependencies .callout-warning}
### Dealing with Dependencies

Everything not included in the project folder that is required
for running the project is called a dependency.
Dependencies are possible breaking points in the future,
therefore it's best to keep them at a minimum.
If you cannot avoid taking a dependency,
make sure that it's available to users of your project in the long term.
In the following, we will discuss two common examples of dependencies:
R packages and downloaded files.

#### R Packages

R packages that you use in your analysis are an obvious example of dependencies.
Their version is recorded by `renv`,
but you also need to ensure that they are available for download.
First, identify the source from which you installed your packages:

```{.r filename="Console"}
# First, install {pak} and {sessioninfo}
renv::install(c(
  "pak",
  "sessioninfo"
))

# Which R packages does the project directly depend on?
deps <- renv::dependencies()$Package |>
  unique()

# Which R packages does the project indirectly depend on?
deps <- deps |>
  pak::pkg_deps(dependencies = NA) |>
  getElement("package")

# Get information about their source
sessioninfo::package_info(deps)
```

If the `source` column only contains the entries `CRAN`, `RSPM`, or `Bioconductor`,
they are already archived.
If the `source` column instead mentions something else (e.g., `GitHub`),
you need to make sure yourself that the package is available to users of your project.
You can either archive the package,
for example in the [Software Heritage archive](https://archive.softwareheritage.org/).
Or you store a copy of the package in the project folder
-- of course, you need to make sure that you are allowed to do so from a copyright perspective
and comply with its license.

#### Downloaded Files

If your code interacts with the internet, for example, to download files,
this is another common dependency with the risk of breaking in the future.
If possible, store a copy of the downloaded file in your project folder.
Alternatively, you can upload it to a permanent repository such as Zenodo
(discussed [later](make_readme.qmd)).
:::

### 2025-03-05 make_readme.qmd
Rework README
Rework the instructions for creating
a README to better fit to papers/
manuscripts, rather than to software
projects.

Having settled on a license, it is time to add a final touch.
Imagine returning to your project in five years,
having forgotten most of the details of what you did exactly.
What would be useful to know in order to quickly understand
what is going on in the project?
This is what needs to be described in the README.
While you could just start writing along,
most READMEs have some sections in common which we describe below.

Name

:   How is the project called?

Badges/Project Status

:   Badges typically report quick facts, like
    [whether the project is actively maintained][repostatus],
    [how many dependencies it has][tinyverse],
    or [whether it has been published in a package repository][metacran].
    Sometimes, the license, any associated DOIs,
    or quality metrics like code coverage by unit tests
    are communicated via badges as well.

[repostatus]: https://www.repostatus.org/
[tinyverse]: https://tinyverse.netlify.app/
[metacran]: https://www.r-pkg.org/services#badges

Description

:   What is the project about? What are its features? Why was it created?

Visuals

:   Is there anything you can show that demonstrates how the project can be used?
    Screenshots or other visuals can make the README more appealing.

Installation/Dependencies

:   What steps need to be taken to run the project? What software needs to be installed?
    R itself and the R packages are already documented as this project uses `renv`.
    Therefore you can focus on all other dependencies,
    such as the system dependencies of R packages
    as well as the version of Quarto.[^renv-quarto]
    Also, don't forget to mention software that you have used for any manual steps.
    See @sec-dependencies for additional information.
    It may make sense to provide these information in the README
    even if you already cite all programs and their version numbers in your manuscript.

@@

Usage

:   Which files does the project folder contain? How are they organized?
    How can one run the project -- is there a master script
    or a particular order in which any scripts need to be executed?
    How long does it take to run all scripts?
    Is there additional documentation available?

Support

:   Do you offer support or help, for example, via GitHub discussions
    or a mailing list?

Contributing

:   If the project is active: Can other people contribute? How?
    Do you accept contributions? Do you review issues?
    This section is sometimes outsourced into a file called `CONTRIBUTING.md`.

Authors

:   Who was involved in creating this project? This involves you,
    your co-authors, and anybody you accepted contributions from.

Citation

:   Is there a recommended way to cite this project?
    Is there a published article associated with it
    that you would like to have cited?

License

:   Under which licenses are the works in this project folder available?

## Installation/Dependencies {#sec-dependencies}

@@

```{.r filename="Console"}
# First, install pak
renv::install("pak")

# Find all R package dependencies
deps <- renv::dependencies()$Package |>
   unique() |>
   pak::pkg_deps(dependencies = NA) |>
   getElement("package")

@@

Of course, all the system dependencies identified until now
may have dependencies on their own. Use your own judgement to decide when not to dig deeper.

Create your README now as the file `README.md`. If you feel stuck,
you can have a look at the following examples:

::: {#tip-name .callout-tip collapse="true"}
### Name

@@

```
:::

::: {#tip-project-status .callout-tip collapse="true"}
### Badges/Project Status

```{.md .code-overflow-wrap filename="README.md"}
[![Project Status: Unsupported – The project has reached a stable, usable state but the author(s) have ceased all work on it. A new maintainer may be desired.](https://www.repostatus.org/badges/latest/unsupported.svg)](https://www.repostatus.org/#unsupported)
```
:::

::: {#tip-description .callout-tip collapse="true"}
### Description

```{.md .code-overflow-wrap filename="README.md"}
This project contains the Quarto manuscript of our study on penguins. It is written in R and uses `renv` to track its dependencies.
```
:::

::: {#tip-installation .callout-tip collapse="true"}
### Installation/Dependencies

``````{.md .code-overflow-wrap filename="README.md"}
## Dependencies

@@

The folder `_extensions` contains the `apaquarto` extension which is used to typeset the PDF accoording to APA guidelines.

@@

>>>>

Having settled on a license, it is time to add a final touch.
Imagine returning to your project in five years,
having forgotten most of the details of what you did exactly.
What would be useful to know in order to quickly understand
what is going on in the project?
This is what needs to be described in the README.
While you could just start writing along,
it is helpful to provide at least the following information in sections on their own.

### Name and Description

How is the project called?
What is it about?
Which files does the project folder contain?
How are they organized?

### Involved Data

Are any (empirical) data involved (e.g., being analyzed or used as input)?
From which sources can they be obtained?
Are they already included in the project folder?
Where is their data dictionary located?
Which terms, usage restrictions, or licenses apply?
If they are not publicly available,
is an alternative, synthesized version provided?

### Computational Requirements

What software needs to be installed to run the analysis
-- in other words, what are its dependencies?
This also includes software that you have used for any manual steps.
If the code has particular hardware requirements
(e.g., in terms of processor or memory),
these should be also noted.
Finally, for steps that take more than a couple of seconds,
the approximate runtime should be indicated.

@nte-dependencies provides more information
on determining the dependencies of an R project.

### Usage

How can one run the project -- is there a master script
or a particular order in which any scripts need to be executed?
Provide detailed instructions for running the full project.

### List of Results

For every result (i.e., number, figure, or table) that is
computed in the project and displayed in the manuscript,
indicate where exactly it is computed.

### Citation

Is there a recommended way to cite this project?
Is there a published article associated with it
that you would like to have cited?

### License

Under which licenses are the works in this project folder available?

Create your README now as the file `README.md`.

::: {#nte-dependencies .callout-note}
### Identifying R Dependencies

R itself and the R packages are already documented as this project uses `renv`.
Therefore you can focus on all other dependencies,
such as the system dependencies of R packages
as well as the version of Quarto.[^renv-quarto]

@@

@@

```{.r filename="Console"}
# Find all R package dependencies
deps <- renv::dependencies()$Package |>
  unique() |>
  pak::pkg_deps(dependencies = NA) |>
  getElement("package")

@@

Of course, all the system dependencies identified until now
may have dependencies on their own.
Use your own judgement to decide when not to dig deeper.
:::

If you feel stuck, you can have a look at the following examples:

::: {#tip-name-description .callout-tip collapse="true"}
### Name and Description

@@

This project contains the Quarto manuscript of our study on penguins ("Manuscript.qmd"). It is written in R and uses `renv` to track its dependencies.

The folder `_extensions` contains the `apaquarto` extension which is used to typeset the PDF accoording to APA guidelines.
```
:::

::: {#tip-data .callout-tip collapse="true"}
### Involved Data

```{.md .code-overflow-wrap filename="README.md"}
## Involved Data

The manuscript analyzes the "palmerpenguins" data set available from <https://cran.r-project.org/package=palmerpenguins>. The data is stored as "data.csv" and documented in the file "data_dictionary.html". It is made available under CC0 1.0.
```
:::

::: {#tip-computational-requirements .callout-tip collapse="true"}
### Computational Requirements

``````{.md .code-overflow-wrap filename="README.md"}
## Computational Requirements

@@

@@

::: {#tip-list-of-results .callout-tip collapse="true"}
### List of Results

```{.md .code-overflow-wrap filename="README.md"}
## List of Results

- In-text numbers in the section "results": Calculated in the chunk "t-test" within "Manuscript.qmd"
- Table 1: Calculated in the chunk "tbl-descriptive-statistics" within "Manuscript.qmd"
- Figure 2: Calculated in the chunk "fig-bill-length-comparison" within "Manuscript.qmd"
```
:::
